Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2020-08-06 paper/notes.tex
Adds code to limit a, to prevent simulation crashes.

On arriving at the tectum, axon growth cones must determine their direction of
movement. The model encapsulates the quantitative hypotheses that growth cone
movement is a combination of diffusion and guidance. The guidance is subject
to sampling error and thus a second source of noise can be introduced to the
model. This noise varies with time during the simulation and can be considered
to have two contributions; one from variations in the true expression level of
the signalling gradient and one from sampling error measuring it. We will
combine these into a single error. \textbf{Or maybe not:} Because the
no-competition model (see Section \ref{sec:nocomp}) has a function
$f(\gamma, \rho)$, which represents the axons' ability to sample the absolute
value of $\rho$, it is important to represent the variability of the true
value of $\rho(\mb{x})$.

@@

\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i,
\end{equation}

the term for the divergence of $\mb{J}$ can be written:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right).
\end{equation}
%
In the barrels model, each $\rho_j(\mb{x})$ was assumed static, allowing the
definition of a time-independent $\mb{g}_i(\mb{x})$:
%
\begin{equation}
\mb{g}_i(\mb{x}) \equiv \sum_{j=1}^{M} \gamma_{i,j} \nabla\rho_j(\mb{x}).
\end{equation}
%
Here, we represent the time-dependence of the signalling molecule expression
by noting that $\rho$ is now a function of time.  The error which the growth
cones make in sampling $\rho$ and estimating $\nabla\rho_j(\mb{x})$ is
$\xi_{\rho'}(t)$:
%
\begin{equation}
\mb{g}_i(\mb{x},t) \equiv \sum_{j=1}^{M} \gamma_{i,j} \big( \nabla\rho_j(\mb{x},t)
+ \xi_{\rho'}(t) \big).
\end{equation}
%
One problem with this representation is that the equations represent a
population model. Within one spatial element of the discretized solution of
the differential equations, the error, $\xi_{\rho'}(t)$ can have only one value at
time $t$; this value is applied to all the individual axon growth cones that
are represented by that element. Nonetheless, it's a useful, and fairly simple
way to represent the guidance error in the system. In practice, we ignore the
variation in $\rho_j$ and write
%
\begin{equation}
\mb{g}_i(\mb{x},t) \equiv \sum_{j=1}^{M} \gamma_{i,j} \big( \nabla\rho_j(\mb{x})
+ \xi_{\rho'}(t) \big).
\end{equation}

@@

Here, we seek to model the `pure' chemospecificity theory in which axons are
guided by orthogonal morphogen gradients until they reach their destination,
which they must be able to sense. The axons, guided by the \emph{gradient} of
the morphogen concentration are assumed also to be able to sample
the \emph{absolute} concentration of the signalling molecules.  A reduction in
branching, and possibly an enhancement of synaptogenesis are triggered when
the concentrations of the two orthogonal morphogens are simultaneously
measured to be within the axon's `stopping' ranges.

To find a way to implement this in a population model similar to the
competitive model, we start from the branching equation (dropping the
competition term, $\chi_i$):

@@

Where $f_i$ is a function which permits branching far away from a given axon's
target region and tends to zero inside the target region. It represents an
axon's internal `stopping process'. Note carefully that this internal axonal
function does \emph{not} depend directly on space; it is a function of the
interaction parameters $\gamma$ and of the signalling molecule density,
$\rho$. Note that $\tilde{\rho}_j(\mb{x})$ is the axon's \emph{measurement} of
$\rho_j(\mb{x})$ and may include measurement noise, $\xi_\rho$:
$\tilde{\rho}_j = \rho_j + \xi_\rho$. Given that the morphogen gradients are
orthogonal, $(\tilde{\rho}_1,\tilde{\rho}_2)$ defines a vector field
$\vec{\rho}\,(\mb{x})$. Similarly, we can combine $(\gamma_{i,1},\gamma_{i,2})$
into the vector $\vec{\gamma}$.

@@

>>>>

On arriving at the tectum, axon growth cones must determine their direction of
movement. The model encapsulates the quantitative hypothesis that growth cone
movement is a combination of diffusion and guidance. The guidance is subject
to sampling error and thus a second source of noise can be introduced to the
model. This noise varies with time during the simulation and can be considered
to have two contributions; one from variations in the true expression level of
the signalling gradient and one from sampling error measuring it. I will
combine these into a single error. \textbf{Or maybe not:} Because the
no-competition model (see Section \ref{sec:nocomp}) has a function
$f(\gamma, \rho)$, which represents the axons' ability to sample the absolute
value of $\rho$, it is important to represent the variability of the true
value of $\rho(\mb{x})$. \textbf{Update (July 2020):} Since I was last
thinking about this, I added length- and amplitude-variable noise to $\rho$,
so I will use that code here, and additionally consider noise in the sampling
of $\rho$.

@@

\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla
a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) + \chi_i \right) - \frac{\partial c_i}{\partial t},
\end{equation}

the term for the divergence of $\mb{J}$ can be written:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) = \nabla\vcdot\left(D \nabla
  a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})+ \chi_i \right).
\end{equation}
%
In the barrels model, each $\rho_j(\mb{x})$ was assumed static, allowing the
definition of a time-independent $\mb{g}_i(\mb{x})$:
%
\begin{equation}
\mb{g}_i(\mb{x}) \equiv \sum_{j=1}^{M} \gamma_{i,j} \nabla\rho_j(\mb{x}).
\end{equation}
%
Here, I represent the time-dependence of the signalling molecule expression
by noting that $\rho$ is now a function of time.  The error which the growth
cones make in sampling $\rho$ and estimating $\nabla\rho_j(\mb{x})$ is
$\xi_{\rho'}(t)$:
%
\begin{equation}
\mb{g}_i(\mb{x},t) \equiv \sum_{j=1}^{M} \gamma_{i,j} \big( \nabla\rho_j(\mb{x},t)
+ \xi_{\rho'}(t) \big).
\end{equation}
%
One problem with this representation is that the equations represent a
population model. Within one spatial element of the discretized solution of
the differential equations, the error, $\xi_{\rho'}(t)$ can have only one value at
time $t$; this value is applied to all the individual axon growth cones that
are represented by that element. Nonetheless, it's a useful (and fairly simple)
way to represent the guidance error in the system. In practice, I ignore the
variation in $\rho_j$ and write
%
\begin{equation}
\mb{g}_i(\mb{x},t) \equiv \sum_{j=1}^{M} \gamma_{i,j} \big( \nabla\rho_j(\mb{x})
+ \xi_{\rho'}(t) \big).
\end{equation}

@@

Here, I seek to model the `pure' chemospecificity theory in which axons are
guided by orthogonal morphogen gradients until they reach their destination,
which they must be able to sense. The axons, guided by the \emph{gradient} of
the morphogen concentration are assumed also to be able to sample
the \emph{absolute} concentration of the signalling molecules.  A reduction in
branching, and possibly an enhancement of synaptogenesis are triggered when
the concentrations of the two orthogonal morphogens are simultaneously
measured to be within the axon's `stopping' ranges.

To find a way to implement this in a population model similar to the
competitive model, I start from the branching equation (dropping the
competition term, $\chi_i$):

@@

Where $f_i$ is a function which permits branching far away from a given axon's
target region and tends to zero inside the target region. It represents an
axon's internal `stopping process'. Note carefully that this internal axonal
function does \emph{not} depend directly on space; it is a function of the
interaction parameters $\gamma$ and of the signalling molecule density,
$\rho$. Note that $\tilde{\rho}_j(\mb{x})$ is the axon's \emph{measurement} of
$\rho_j(\mb{x})$ and may include measurement noise, $\xi_\rho$:
$\tilde{\rho}_j = \rho_j + \xi_\rho$. Given that the morphogen gradients are
orthogonal, $(\tilde{\rho}_1,\tilde{\rho}_2)$ defines a vector field
$\vec{\rho}\,(\mb{x})$. Similarly, $(\gamma_{i,1},\gamma_{i,2})$ can be
combined to give the vector $\vec{\gamma}$.

@@

\section{Implementation notes}

\subsection{Stability}

The model can become unstable due to the unbounded nature of $a$. Essentially,
$a$ can blow up in the computation of $\dvrg J$.

Stability with respect to $D$ is a particular issue, as is stability in the
presence of noise.

\subsection{Stability for small $D$}

In the basic, $N=56$ wild type simulation, $D$ seems to need to be close to
0.02. Unfortunately, the simulation is unstable for $\delta t=0.0001$ and
$d=0.015$. Having fixed a bug which prevented me from reducing $d$ very much,
and having implemented fixed seed random numbers for repeatable sims,
I can now try to determine if a smaller $d$ will help the simujlation compute,
though note that as $d$ is reduced, so $\delta t$ must be reduced, leading to
longer simulation computations.

Note: $\epsilon=1.2$ here in the following:

D=0.02, dt=0.00005, d=0.007: Crash at step 13803.

D=0.01, dt=0.0001, d=0.015: Crash at step 2629.

D=0.015, dt=0.0001, d=0.015: Crash at step 4864.

D=0.02, dt=0.0001, d=0.015: Crash at step 7280.

D=0.025, dt=0.0001, d=0.015: Crash at step 9649.

D=0.03, dt=0.0001, d=0.015: Crash at step 12394.

This pattern is similar to that shown in Fig.~2 of the Barrels paper. To
enable $D$ to be low, it's necessary to reduce $\epsilon$, BUT I want to
explore increased competition, whilst maintaining low $D$.

Thus, some sort of cap on $a$ may be called for.

But, lets first see how much reduction of $\epsilon$ is required. Starting
from D=0.01, dt=0.0001, d=0.015: Crash at step 2629, with $\epsilon=1.2$,
reduce $\epsilon$.

$\epsilon=0.8$: Crash step 4779.

$\epsilon=0.4$: Crash step 9802.

$\epsilon=0.1$: Computation makes it to step 24000 with no crash, but that's
not enough competition for an even result.

\subsection{Bounding $a$}

I've already implemented systems in which $a$ is bounded. Simplest is probably
best - a piecewise linear transfer function on $a$. Thus, I've added the code
to rd_barrel.h to compute a transformed version of $a$ called $\bar{a}$ and
use this in the computation. It mitigates the instabilities without
making much change to the operation of the model. It adds an additional
parameter to the model, $a_{max}$, which I've set to 0.5 by default.

\subsection{Stability for noise}

Write me.
